Exploiting First-Class Arrays in Fortran for Accelerator Programming

### 2010-12-15 opencl/papers/programming-model.tex
Added function discussion.

Pure procedures, like elemental functions, must be free of side effects.
(athough it may change the value of array elements in an array argument,
unlike elemental arguments which must have the intent(in) attribute).
The absence of side effects removes ordering constraints from the compiler
allowing it to invoke pure functions out of order.  Procedures and functions
of this sort are also common in pure functional languages like Haskell,
and are exploited by compilers in order to emit parallel code automatically
due to their suitability for compiler-level analysis.

Since pure procedures don't have side effects they a candidates for running on
accelerators in OpenCL.  Currently ForOpenCL transforms pure procedures to
OpenCL kernels that emph{do not} call other procedures, except for elemental
functions.

@@

Regions are similar to the shift operator as they can be used to reference
portions of the array that are shifted with respect to the interior portion.
However, unlike the shift operator, regions are not expressed in terms of
boundary conditions and thus don't explicitly \emph{require} a knowledge of
not the application of boundary conditions locally.  Thus, as will be shown
below, regions are more suitable for usage by OpenCL thread groups which
access only local subsections of an array stored in global memory.

Unfortunately, as regions functions return pointers to arrays, they cannot
be used in pure procedures in current Fortran.  We relax this constraint in
ForOpenCL for region pointers only.

\subsection{New Functions}

% TODO - Craig - describe these functions
%   region
%   interior (can be written in terms of regions) but also specifies the thread
%       group domain)
%   more ...
%

@@

% TODO - Matt - do you want a try at this.

@@

\begin{itemize}
	\item There are no loops or index variables to keep track of.  Off by one
index errors are a common programming mistake.
        \item The written code is closer to the algorithm, easier to understand, and is usually substantially shorter.
	\item Semantically the intrinsic functions return arrays by value (though not regions).  This is usually what the algorithm requires.
	\item Because pure and elemental function are free from side effects, it is easier for a compiler to schedule the work to be done in parallel.
\end{itemize}

@@

>>>>

Pure procedures, like elemental functions, must be free of side effects.
(athough they may change the value of array elements in an array argument,
unlike elemental arguments which must have the intent(in) attribute).
The absence of side effects removes ordering constraints from the compiler
allowing it to invoke pure functions out of order.  Procedures and functions
of this sort are also common in pure functional languages like Haskell,
and are exploited by compilers in order to emit parallel code automatically
due to their suitability for compiler-level analysis.

Since pure procedures don't have side effects they a candidates for running on
accelerators in OpenCL.  Currently ForOpenCL transforms pure procedures to
OpenCL kernels that \emph{do not} call other procedures, except for elemental
functions.

@@

Regions are similar to the shift operator as they can be used to reference
portions of the array that are shifted with respect to the interior portion.
However, unlike the shift operator, regions are not expressed in terms of
boundary conditions and thus don't explicitly \emph{require} a knowledge of
nor the application of boundary conditions locally.  Thus, as will be shown
below, regions are more suitable for usage by OpenCL thread groups which
access only local subsections of an array stored in global memory.

\subsection{New Procedures}

Unlike OpenMP, we require limited use of compiler directives, although
two are used to enforce the semantics required by the programming
model.  These directives are:
\begin{itemize}
   \item !$OFP CONTIGUOUS: specifies that an dummy array variable is
   contiguous in memory. CONTIGUOUS is an attribute in Fortran 2008.
   \item !$OFP KERNEL: specifies that a pure subroutine can be
     transformed to an OpenCL kernel.
\end{itemize}

Three new procedures (in addition to the intrinsic shift function) are
defined in Fortran that are used in array-syntax operations.  Each
procedure takes a integer array halo argument that specifies the
number of ghost cells on either side of a region, for each dimension.
For example {\tt halo = [left, right, down, up]} specifies a halo for
a two-dimensional region.  These functions are:
\begin{itemize}
   \item transfer_halo(array, halo): an impure subroutine that
     exchanges halo regions between nodes using MPI or Fortran
     coarrays. Array is the memory stored on the node representing the
     node-local portion of a virtual global array. Not used in this work.
   \item interior(array, halo): a pure function that returns a copy of
     the interior portion of the array specified by halo. Array is
     local to a node.
   \item region(array, halo): a pure function that returns a copy of
     the portion of the array specified by halo.  Array is local to a node.
\end{itemize}

It should be noted that the two functions {\tt interior} and {\tt
  region} are pure and thus can be called from within a pure kernel
procedure.  These two functions are part of the language recognized by
the compiler and though the two functions return a copy of a portion
of an array \emph{semantically}, the compiler is not forced to
actually make a copy and is free to enforce copy semantics through
other means.

@@

% TODO - Matt - do you want a try at this.  Not sure we have room for
% this subsection. I briefly mention node parallelism in other sections.

@@

\begin{itemize}
	\item There are no loops or index variables to keep track of.  Off by one
index errors are a common programming mistake.
        \item The written code is closer to the algorithm, easier to understand, and is usually substantially shorter.
	\item Semantically the intrinsic functions return arrays by value.  This is usually what the algorithm requires.
	\item Because pure and elemental function are free from side effects, it is easier for a compiler to schedule the work to be done in parallel.
\end{itemize}

@@

A kernel procedure (specified by the !$OFP KERNEL directive) must be
pure and all array variables must be declared as contiguous.  A kernel
procedure may not call other procedures except for limited intrisic
functions (primarily math), user-defined elemental function, and the
{\tt interior} and {\tt region} functions.
